---
title: "Recitation for Week05"
author: "Prateek Jain"
format: revealjs
editor: visual
execute: 
  echo: true
editor_options: 
  chunk_output_type: console
---

## Introduction

```{=html}
<script type="text/x-mathjax-config">
MathJax.Hub.Register.StartupHook("TeX Jax Ready",function () {
  MathJax.Hub.Insert(MathJax.InputJax.TeX.Definitions.macros,{
    cancel: ["Extension","cancel"],
    bcancel: ["Extension","cancel"],
    xcancel: ["Extension","cancel"],
    cancelto: ["Extension","cancel"]
  });
});
</script>
```
-   We are going to manipulate the Pareto Distribution, which is often used to model skewed, positive RVs, such as income

-   But think about the dollar price of Bitcoin in $50$ years

-   Suppose that it is uniformly distributed between $0$ and $\theta$, where $\theta$ is unknown but you have Pareto beliefs about $\theta$

-   Focus on the Bayesian mathematical manipulations to estimate $\theta$, evaluate hypotheses, and predict the future

-   Create a blank .R file to enter your code

-   `source("pareto.R")` from Week05 to define `rpareto`

## Cumulative Density Function (CDF)

-   Let $\Theta = \left[k,\infty\right)$ and $k,w > 0$. The CDF of the Pareto distribution is $$
    F\left(\theta \mid k,w\right) = 1 - \left(\frac{k}{\theta}\right)^w
    $$ if $\theta \geq k$ or else $F\left(\theta \mid k,w\right) = 0$.

-   How can we define `ppareto` in R?

. . .

```{r}
ppareto <- function(x, k, w) ifelse(x >= k, 1 - (k / x)^w, 0)
```

```{r}
k <- 65000
w <- 1.5 # you could choose other positive values
```

## Probability Density Function (PDF)

-   If the CDF is $F\left(\theta \mid k,w\right) = 1 - \left(\frac{k}{\theta}\right)^w$ , what is the PDF?

. . .

$$
\frac{\partial F\left(\theta \mid k,w\right)}{\partial \theta} = -w\left(\frac{k}{\theta}\right)^{w - 1} \times \frac{-k}{\theta^2} = \frac{wk^w}{\theta^{w + 1}} = f\left(\theta \mid k,w\right)
$$

-   If you cannot work that out for yourself, you could type `differentiate 1 - (k / x)^w` at <https://www.wolframalpha.com/> (which also does integrals)

-   How would we implement the PDF in R?

. . .

```{r}
dpareto <- function(x, k, w) ifelse(x >= k, w * k^w / x^(w + 1), 0)
```

## Graph of PDF

```{r}
#| echo: false
library(ggplot2)
ggplot() +
  xlim(0, 500000) +
  geom_function(fun = dpareto, args = list(k = k, w = w)) +
  labs(x = "theta",
       y = "density")
source("pareto.R")
```

## Validation

-   How could we check that `dpareto` is correct?

. . .

```{r}
integrate(dpareto, lower = k, upper = Inf, k = k, w = w)
```

. . .

-   If `dpareto` & `rpareto` are correct, what is going on here?

```{r}
rbind(first  = mean(rpareto(10^7, k, w)),
      second = mean(rpareto(10^7, k, w)),
      third  = mean(rpareto(10^7, k, w)),
      fourth = mean(rpareto(10^7, k, w)),
      fifth  = mean(rpareto(10^7, k, w)),
      sixth  = mean(rpareto(10^7, k, w)))
```

## Expectation

-   What is an expression for $\mathbb{E}\theta \mid k,w$?

. . .

$$\mathbb{E}\theta = \int_{k}^\infty \theta f\left(\theta \mid k, w\right) d\theta =
  \int_{k}^\infty \theta \frac{w k^w}{\theta^{w + 1}} d\theta = \\
  \int_{k}^\infty w \left(\frac{k}{\theta}\right)^{w} d\theta = 
  w k^{w} \int_{k}^\infty \theta^{-w} d\theta = 
  \left.\frac{w k^{w}}{1 - w} \theta^{1 -w}\right|_{k}^\infty$$If $w \leq 1$, then this definite integral is *infinite*. If $w > 1$, then the expectation of $\theta$ is finite, namely $\mu = \frac{w k^{w}}{w - 1} k^{1 -w} = \frac{w k}{w - 1}$.

-   Can type `integrate w * (k / x)^w from x = k to infinity` at <https://www.wolframalpha.com/>

## Prior Predictive Density Function

-   Suppose $X$ (e.g. the dollar price of Bitcoin in the future) is uniform between $0$ and $\theta$ with PDF $f\left(x \mid \theta\right) = \frac{1}{\theta}$

-   What is the PDF of $X$ irrespective of $\theta$, if our beliefs about $\theta$ are distributed Pareto with minimum $k$ and shape $w$?

. . .

$$f\left(x \mid k, w\right) = 
f\left(x \bigcap \bcancel{\theta} \mid k, w\right) =
\int_{\max\left(k,x\right)}^{\infty} \frac{1}{\theta} 
\frac{w k^w}{\theta^{w + 1}} d\theta = \\
w k^w \int_{\max\left(k,x\right)}^{\infty} \theta^{-w - 2} d\theta = 
\left.w k^w\frac{\theta^{-w - 1}}{-w - 1}\right|_{\max\left(k,x\right)}^\infty = \\
w k^w \max\left(k,x\right)^{-w - 1} / \left(w + 1\right)$$

## Drawing Joint Random Values

-   In R, draw from the joint distribution of $\theta \bigcap X \mid k,w$

. . .

```{r}
#| message: false
library(dplyr)
draws <- tibble(theta = rpareto(10^7, k, w),
                x = runif(10^7, min = 0, max = theta))
```

-   What is the prior probability of the hypothesis that $X > 100,000$?

. . .

```{r}
summarize(draws, prob = mean(x > 100000))
```

## Posterior PDF

-   Suppose you observe $x$ (i.e. the dollar price of Bitcoin today)

-   What is the posterior PDF for $\theta \mid k, w, x$?

. . .

$$f\left(\theta \mid k, w, x\right) =
\frac{ w k^w / \theta^{w + 1} \times 1 / \theta }
{w k^w \max\left(k,x\right)^{-w - 1} / \left(w + 1\right)} = 
\frac{w^\ast \left(k^\ast\right)^{w^\ast}}{\theta^{w^\ast + 1}}$$

where $w^\ast = w + 1$ and $k^\ast = \max\left(k,x\right)$

-   The posterior distribution remains in the Pareto family and the Pareto prior is conjugate with the uniform likelihood
-   Any term in the numerator that doesn't depend on $\theta$ cancels

## Posterior PDF Given $N$ Observations

-   Suppose you have $N$ observations

-   What is the posterior PDF of $\theta \mid k, w, x_1, x_2, \dots, x_N$?

. . .

-   The posterior PDF conditional on $N$ observations can be obtained by updating the Pareto prior $N$ times, yielding $$f\left(\theta \mid k, w, x_1, x_2, \dots, x_N\right) =
    \frac{w^\ast \left(k^\ast\right)^{w^\ast}}{\theta^{w^\ast + 1}}$$ where $w^\ast = w + N$ and $k^\ast = \max\left(k, x_1, x_2, \dots, x_N\right)$

## Posterior Predictive PDF {.smaller}

-   How would you predict $X$ in $50$ years?

. . .

-   The posterior predictive density for a future $x_{N + 1}$ is in the same form as the prior predictive density found earlier (and can be derived in the same way), except evaluated at $k^\ast$ and $w^\ast$, rather than $k$ and $w$: $$f\left(x_{N + 1} \mid w, k, x_1, x_2, \dots, x_N\right) = 
    f\left(x_{N + 1} \bigcap \bcancel{\theta} \mid w, k, x_1, x_2, \dots, x_N\right) = \\
    \int_k^\infty f\left(x_{N + 1} \mid \theta\right) 
    f\left(\theta \mid x_1, x_2, \dots, x_N, k, w\right)d\theta = \\
    \int_{\max\left(k^\ast, x_{N + 1}\right)}^\infty f\left(x_{N + 1} \bigcap \theta \mid 
    k^\ast, w^\ast\right)d\theta =
    w^\ast \left(k^\ast\right)^{w^\ast}
    \frac{\max\left(k^\ast,x_{N + 1}\right)^{-w^\ast - 1}}{w^\ast + 1}$$

-   After you beliefs about $\theta$ have been updated by conditioning on past data, you do not explicitly need the past data any more; i.e. $k^\ast$ and $w^\ast$ are all you need to predict $x_{N + 1}$

## Conclusion

-   You need a model, which includes your prior beliefs about the unknown parameters as expressed through a PDF

-   In this case, the prior predictive PDF is elementary so we can derive the posterior PDF, which is also in the Pareto family

-   From there, we can draw from the posterior predictive distribution of future data

-   We can evaluate hypotheses about $\theta \mid k,w,x$ or $x_{n + 1} \mid k, w, x_1, x_2, \dots, x_N$

-   But usually we will have to do it all with draws
